---
title: "Death Rate from Heart Disease and Stroke in the USA - CDC data"
author: "Gabriel Falcao Alencar"
date: "January, 2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Heart disease is the leading cause of death in the USA, accounting for approximately 1 in every 4 deaths (@ref). **I will write more about introduction and include some figures in the next few days**.

## Prerequisites

To perform these maps, you will need to have several extra packages. I'm not going into how to install them, but a quick google search should tell you how to. The main libraries that we will need are:

1. [tidyverse](https://www.tidyverse.org/): this is a package created by Hadley Wickham. This "umbrella" package was designed to make data science analysis and visualition faster and easier. The main packages that we will use in this analysis are:
    + [dplyr](http://dplyr.tidyverse.org/): grammar of data manipulation, i.e. how to import, select, filter, and arrange your data for analysis;
    + [ggplot2](http://ggplot2.tidyverse.org/): system for declaratively creating graphics, i.e. you provide the data, tell it how to map, and it takes care of details;
    + [stringr](http://stringr.tidyverse.org/): provides a set of functions designed to make working with strings as easy as possible.
2. [maps](https://cran.r-project.org/web/packages/maps/README.html): package for displaying maps;
3. [mapdata](https://www.rdocumentation.org/packages/mapdata/versions/2.2-6): supplements the `maps` package, providing some larger and/or higher-resolution databases.
4. [ggmap](https://github.com/dkahle/ggmap): package to retrieve [raster map tiles](https://en.wikipedia.org/wiki/Tiled_web_map) (map displayed by joining dozens of individually requested image files over the internet) from popular online mapping service like Google Maps, and others.

## Load the libraries
```{r loading-packages}
library(ggplot2)
library(maptools)
library(ggalt)
library(ggthemes)
library(albersusa) # devtools::install_github("hrbrmstr/albersusa")
library(rgeos)
library(viridis)
library(scales)
library(tidyverse)
```

## Getting data about counties

Data was obtained from the [CDC](https://www.cdc.gov/). For this first analysis, I downloaded the data for **Deaths - Total Cardiovascular Disease** for both genders, 35+ year old individuals, from 2005 to 2015 (two year increment, 2005-2007, 2007-2009, etc). The files also have risk factors associated with CVD, such as **Obesity**, **Diabete**, and **Exercise**.

```{r loading-data}
#r stands for raw right now
r.total.05.07 <- as.data.frame(read.table("data/2005-2007-cdc-report-heart-disease-and-risk-factors.txt", header=TRUE, sep=",", colClasses = 'character'))
r.total.07.09 <- as.data.frame(read.table("data/2007-2009-cdc-report-heart-disease-and-risk-factors.txt", header=TRUE, sep=",", colClasses = 'character'))
r.total.09.11 <- as.data.frame(read.table("data/2009-2011-cdc-report-heart-disease-and-risk-factors.txt", header=TRUE, sep=",", colClasses = 'character'))
r.total.11.13 <- as.data.frame(read.table("data/2011-2013-cdc-report-heart-disease-and-risk-factors.txt", header=TRUE, sep=",", colClasses = 'character'))
r.total.13.15 <- as.data.frame(read.table("data/2013-2015-cdc-report-heart-disease-and-risk-factors.txt", header=TRUE, sep=",", colClasses = 'character'))
head(r.total.05.07)
```

So, there are several things that we need to change. First, we will select the columns that we want to look at the moment, in our case, `cnty_fips` and `Value`. We will then proceed to rename into `fips` and `avgdeath`. And finally, we will change the `-1` value to `NA`. That is all that we need to start the plotting of our data.

```{r}
# Gonna do the 05-07 first to see how it goes. After that do the rest
heart05 <- select(r.total.05.07, cnty_fips, Value) %>%
  rename(fips=cnty_fips, avgdeath=Value)

heart05$avgdeath <- as.numeric(replace(heart05$avgdeath, heart05$avgdeath<0, NA))
heart05$fill <- cut(heart05$avgdeath, c(0,300, 400, 500, 600, 2000))
levels(heart05$fill) = c("0-299.9","300-399.1","400-499.1","500-599.1",">600")
head(heart05)
summary(heart05)
```

In theory, that is all we need to start creating our map.

```{r}
cmap <- fortify(counties_composite(), region="fips")
head(cmap)
```

Ok, now we are finally ready to plot the mortality rates

```{r plot-counties}
ggcounties <- ggplot()
ggcounties <- ggcounties + geom_map(data=cmap, map=cmap,
                    aes(x=long, y=lat, map_id=id),
                    color="#2b2b2b", size=0.05, fill=NA)
ggcounties
```


```{r 2005}
gg05 <- ggcounties + geom_map(data=heart05, map=cmap,
                    aes(fill=fill, map_id=fips),
                    color="#2b2b2b", size=0.05)
gg05 <- gg05 + scale_fill_manual(values=c("#10052e","#4a126b","#a52c60","#ed6925","#f7d03c"))
gg05 <- gg05 + theme_void() + theme(legend.position="none")
gg05
```

```{r 2007}
# Gonna do the 05-07 first to see how it goes. After that do the rest
heart07 <- select(r.total.07.09, cnty_fips, Value) %>%
  rename(fips=cnty_fips, avgdeath=Value)

heart07$avgdeath <- as.numeric(replace(heart07$avgdeath, heart07$avgdeath<0, NA))
heart07$fill <- cut(heart07$avgdeath, c(0,300, 400, 500, 600, 2000))
levels(heart07$fill) = c("0-299.9","300-399.1","400-499.1","500-599.1",">600")
head(heart07)
summary(heart07)
gg07 <- ggcounties + geom_map(data=heart07, map=cmap,
                    aes(fill=fill, map_id=fips),
                    color="#2b2b2b", size=0.05)
gg07 <- gg07 + scale_fill_manual(values=c("#10052e","#4a126b","#a52c60","#ed6925","#f7d03c"))
gg07 <- gg07 + theme_void() + theme(legend.position="none")
gg07
```
```{r 2009}
heart09 <- select(r.total.09.11, cnty_fips, Value) %>%
  rename(fips=cnty_fips, avgdeath=Value)

heart09$avgdeath <- as.numeric(replace(heart09$avgdeath, heart09$avgdeath<0, NA))
heart09$fill <- cut(heart09$avgdeath, c(0,300, 400, 500, 600, 2000))
levels(heart09$fill) = c("0-299.9","300-399.1","400-499.1","500-599.1",">600")
head(heart09)
summary(heart09)
gg09 <- ggcounties + geom_map(data=heart09, map=cmap,
                    aes(fill=fill, map_id=fips),
                    color="#2b2b2b", size=0.05)
gg09 <- gg09 + scale_fill_manual(values=c("#10052e","#4a126b","#a52c60","#ed6925","#f7d03c"))
gg09 <- gg09 + theme_void() + theme(legend.position="none")
gg09
```

```{r 2011}
heart11 <- select(r.total.11.13, cnty_fips, Value) %>%
  rename(fips=cnty_fips, avgdeath=Value)

heart11$avgdeath <- as.numeric(replace(heart11$avgdeath, heart11$avgdeath<0, NA))
heart11$fill <- cut(heart11$avgdeath, c(0,300, 400, 500, 600, 2000))
levels(heart11$fill) = c("0-299.9","300-399.1","400-499.1","500-599.1",">600")
head(heart11)
summary(heart11)
gg11 <- ggcounties + geom_map(data=heart11, map=cmap,
                    aes(fill=fill, map_id=fips),
                    color="#2b2b2b", size=0.05)
gg11 <- gg11 + scale_fill_manual(values=c("#10052e","#4a126b","#a52c60","#ed6925","#f7d03c"))
gg11 <- gg11 + theme_void() + theme(legend.position="none")
gg11
```

```{r 2013}
heart13 <- select(r.total.13.15, cnty_fips, Value) %>%
  rename(fips=cnty_fips, avgdeath=Value)

heart13$avgdeath <- as.numeric(replace(heart13$avgdeath, heart13$avgdeath<0, NA))
heart13$fill <- cut(heart13$avgdeath, c(0,300, 400, 500, 600, 2000))
levels(heart13$fill) = c("0-299.9","300-399.1","400-499.1","500-599.1",">600")
head(heart13)
summary(heart13)
gg13 <- ggcounties + geom_map(data=heart13, map=cmap,
                    aes(fill=fill, map_id=fips),
                    color="#2b2b2b", size=0.05)
gg13 <- gg13 + scale_fill_manual(values=c("#10052e","#4a126b","#a52c60","#ed6925","#f7d03c"))
gg13 <- gg13 + theme_void() + theme(legend.position="none")
gg13
```